Identifying and correcting multiple sources of misspecification\\in GWAS summary statistics for prediction

### 2021-08-13 paper/paper-misspec.tex
almost finished with the methods

When using the expected genotypes from imputation (dosages), their standard deviations are usually lower than the expected value under Hardy-Weinberg equilibrium (cf.\ Methods section ``Standard deviations and imputation measures'').
In simulations (cf.\ Methods section ``Data for simulations''), we show that $\text{sd}(G_j)^{true} \approx \text{sd}(G_j)^{imp} / \sqrt{\text{INFO}_j}$, $\text{se}(\hat{\beta}_j)^{true} \approx \text{se}(\hat{\beta}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$, and $\hat{\beta}_j^{true} \approx \hat{\beta}_j^{imp} \cdot \sqrt{\text{INFO}_j}$ (Figures \ref{fig:compare-sd}, \ref{fig:compare-beta}, and \ref{fig:compare-beta-se}).
This is the first correction of summary statistics we try in the simulations below. 
Note that we recompute INFO scores for the subset of 362,307 European individuals used in this paper as they can differ substantially from the ones reported by the UK Biobank for the whole data (Figure \ref{fig:compare-info}).
As a second option, instead of using dosages to perform the GWAS, it has been argued that using multiple imputation (MI) would be more appropriate \cite[]{Palmer2016}.
In simulations, we show that $Z_j^{MI} \approx Z_j^{imp} \cdot \text{INFO}_j$ and $\hat{\beta}_j^{MI} \approx \hat{\beta}_j^{imp} \cdot \text{INFO}_j$ (Figure \ref{fig:compare-mi}).
This is the second correction of summary statistics we try in the simulations below, along with $n_j = N \cdot \text{INFO}_j$. 
Finally, we try an in-between solution as a third correction, using $\hat{\beta}_j^{imp} \cdot \text{INFO}_j$, $\text{se}(\hat{\beta}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$, and $N \cdot \text{INFO}_j$.

@@

\begin{figure}[h]
	\centerline{\includegraphics[width=0.95\textwidth]{simu-info}}
	\caption{Results for the simulations using dosage data. Correction ``sqrt\_info'' corresponds to using $\hat{\beta}_j^{imp} \cdot \sqrt{\text{INFO}_j}$ and $\text{se}(\hat{\beta}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$.
	Correction ``info'' corresponds to using $\hat{\beta}_j^{imp} \cdot \text{INFO}_j$  and $N \cdot \text{INFO}_j$.}
	\label{fig:simu-info}
\end{figure}

@@

Therefore, having the INFO score enables both to correct for it in the QC step to be able to pinpoint possible other problems, and possibly to correct for it and improve prediction as shown in the simulations before.
We try this for the two breast cancer summary statistics. We first compare the standard QC proposed in \cite{prive2020ldpred2} (which ends up filtering on MAF here, which we call ``qc1''). We then also filter out the two long-range LD regions of chromosome 6 and 8 for the OncoArray summary statistics and remove around 500 variants when filtering on differences of MAFs between summary statistics and the validation dataset.
As for the correction for the INFO scores, we use $\hat{\beta}_j^{imp} \cdot \text{INFO}_j$ and $N \cdot \text{INFO}_j$ as we have shown to perform well in simulations.
Although results for both QC used are very similar, correcting for the INFO score really improves prediction, especially for LDpred2-auto [TODO: NUMBERS] (Figure \ref{fig:res_brca}).

@@

Among genetic variants with a minor allele frequency larger than 0.01 and an imputation INFO score larger than 0.4 (as reported in the UK Biobank), we sample 40,000 of them according to the inverse of the INFO score density so that they have varying levels of imputation accuracy (Figure \ref{fig:hist-info}).
We read the UK Biobank data into two different datasets using function \texttt{snp\_readBGEN} from R package bigsnpr \cite[]{prive2017efficient}, one by reading the BGEN data at random according to imputation probabilities, and another one reading it as dosages (i.e.\ expected values according to imputation probabilities).
The first dataset is used as what could be the real genotype calls and the second dataset as what would be its imputed version; this technique was used in \cite{prive2019making}.

@@

We also use the UK Biobank data, and use the same individuals as described in the previous subsection. We sample 10,000 individuals to form a validation set and use the remaining 352,307 individuals as test set.
We restrict to the genetic variants to the 1,054,315 variants used in the LD reference provided in \cite{prive2020ldpred2}.

@@

In this paper, we extensively use the following formula
\begin{equation}\label{eq:approx-sd-lin}
\text{sd}(G_j) \approx \dfrac{\text{sd}(y)}{\sqrt{n_j ~ \text{se}(\hat{\beta}_j)^2 + \hat{\beta}_j^2}} ~,
\end{equation}
where $\hat{\beta}_j$ is the marginal (GWAS) effect of variant $j$, $n_j$ is the GWAS sample size associated with variant $j$, $y$ is the vector of phenotypes and ${G}_j$ is the vector of genotypes for variant $j$.
This formula is used in LDpred2 \cite{prive2020ldpred2,prive2021high}. Note that, for a binary trait for which logistic regression is used, we have instead
\begin{equation}\label{eq:approx-sd-log}
\text{sd}(G_j) \approx \dfrac{2}{\sqrt{n_j^\text{eff} ~ \text{se}(\hat{\beta}_j)^2 + \hat{\beta}_j^2}} ~,
\end{equation}
where $n_j^\text{eff} = \frac{4}{1 / n_j^\text{cases} + 1 / n_j^\text{controls}}$.

We can then impute $n_j$ from equation \eqref{eq:approx-sd-lin} using
\begin{equation}\label{eq:imputeN}
n_j \approx \dfrac{\text{var}(y) / \text{var}(G_j) - \hat{\beta}_j^2}{\text{se}(\hat{\beta}_j)^2} ~.
\end{equation}
In practice, we also bound this estimate to be between $0.5 \cdot N$ and $1.1 \cdot N$, where $N$ is the global sample size.

\subsection*{Standard deviations and imputation measures}

The MACH $\hat{r}_j^2$ measure for variant $j$ is given by $\hat{r}_j^2 = \frac{var(G_j)}{2 \hat{\theta} (1 - \hat{\theta})}$, where $\hat{\theta}$ is the estimated allele frequency \cite[]{marchini2010genotype}.
This measure is also highly concordant with the IMPUTE INFO measure \cite[]{marchini2010genotype}.
Therefore $\text{INFO}_j \approx \frac{var(G_j)}{2 \hat{\theta} (1 - \hat{\theta})}$.

@@

Instead of using a regularized version of the correlation matrix $R$ parameterized by $s$, $R_s = (1 - s) R + s I$, we use $R_{\delta} = R + \delta I$, which makes it clearer that lassosum is also using L2-regularization (therefore elastic-net). 
Then, from \cite{mak2017polygenic}, the solution from lassosum can be obtained by iteratively updating 
\[
\beta_j^{(t)} =
\begin{cases}
\text{sign}\left(u_j^{(t)}\right) \left(\left|u_j^{(t)}\right| - \lambda\right) / \left(\widetilde{X}_j^T \widetilde{X}_j + \delta\right) & \text{if } \left|u_j^{(t)}\right| > \lambda ~, \\
0 & \text{otherwise.}
\end{cases}
\]
where 
\[
u_j^{(t)} = r_j - \widetilde{X}_j^T \left( \widetilde{X} \beta^{(t-1)} - \widetilde{X}_j \beta_j^{(t-1)} \right) ~.
\]
Following notations from \cite{prive2020ldpred2} and denoting $\widetilde{X} = \frac{1}{\sqrt{n-1}} C_n G S^{-1}$, where $G$ is the genotype matrix, $C_n$ is the centering matrix and $S$ is the diagonal matrix of standard deviations of the columns of $G$.
Then $\widetilde{X}_j^T \widetilde{X} = R_{j,.} = R_{.,j}^T$, $\widetilde{X}_j^T \widetilde{X}_j = 1$ and
\[
u_j^{(t)} = \beta_j^{(t-1)} + \widehat{\beta}_j - R_{.,j}^T \beta^{(t-1)} ~,
\]
where $r_j = \widehat{\beta}_j =  \dfrac{\widehat{\beta}_j}{\sqrt{n_j ~ \text{se}(\widehat{\beta}_j)^2 + \widehat{\beta}_j^2}}$ and $\widehat{\beta}_j$ is the GWAS effect of variant $j$ and $n$ is the GWAS sample size \cite[]{mak2017polygenic,prive2021high}.
Then the most time-consuming part is computing $R_{.,j}^T \beta^{(t-1)}$.
To make this faster, instead of computing $R_{.,j}^T \beta^{(t-1)}$ at each iteration ($j$ and $t$), we can start with a vector with only 0s initially (for all $j$) since $\beta^{(0)} \equiv 0$, and then updating this vector when $\beta_j^{(t)} \neq \beta_j^{(t-1)}$ only. Note that only positions $k$ for which $R_{k,j} \neq 0$ must be updated in this vector $R_{.,j}^T \beta^{(t-1)}$.

In this new implementation of the lassosum model, the input parameters are the correlation matrix $R$, the GWAS summary statistics ($\widehat{\beta}_j$, $\text{se}(\widehat{\beta}_j)$ and $n_j$), and the two hyper-parameters $\lambda$ and $\delta$. 
Therefore, except for the hyper-parameters, this is the exact same input as for LDpred2 \cite[]{prive2020ldpred2}.
We try $\delta \in \{0.001, 0.005, 0.02, 0.1, 0.6, 3\}$ by default in lassosum2, instead of $s \in \{0.2, 0.5, 0.8, 1.0\}$ in lassosum.
For $\lambda$, the default in lassosum uses a sequence of 20 values equally spaced on a log scale between 0.1 and 0.001; we now use a sequence between $\lambda_0$ and $\lambda_0 / 100$ by default in lassosum2, where $\lambda_0 = \max_j \left|\widehat{\beta}_j\right|$ is the minimum $\lambda$ for which no variable enters the model because the L1-regularization is too strong.
Note that we do not provide an ``auto'' version using pseudo-validation (as in \cite{mak2017polygenic}) as we have not found it to be very robust (Figure \ref{fig:pseudoval}).
Also note that, as in LDpred2, we run lassosum2 genome-wide using a sparse correlation matrix which assumes that variants further away than 3 cM are not correlated, and therefore we do not require splitting the genome into independent LD blocks anymore (as is required in lassosum).

@@

Authors thank GenomeDK and Aarhus University for providing computational resources and support that contributed to these research results.
This research has been conducted using the UK Biobank Resource under Application Number 58024.

>>>>

The MACH $\hat{r}_j^2$ measure for variant $j$ is given by $\hat{r}_j^2 = \frac{var(G_j)}{2 \hat{\theta} (1 - \hat{\theta})}$, where $\hat{\theta}$ is the estimated allele frequency \cite[]{marchini2010genotype}.
This measure is also highly concordant with the IMPUTE INFO measure \cite[]{marchini2010genotype}.
Therefore, when using the expected genotypes from imputation (dosages), their standard deviations are usually lower than the expected value under Hardy-Weinberg equilibrium because $\text{INFO}_j \approx \frac{var(G_j)}{2 \hat{\theta} (1 - \hat{\theta})}$.
In simulations (cf.\ Methods section ``Data for simulations''), we show that $\text{sd}(G_j)^{true} \approx \text{sd}(G_j)^{imp} / \sqrt{\text{INFO}_j}$, $\text{se}(\hat{\gamma}_j)^{true} \approx \text{se}(\hat{\gamma}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$, and $\hat{\gamma}_j^{true} \approx \hat{\gamma}_j^{imp} \cdot \sqrt{\text{INFO}_j}$ (Figures \ref{fig:compare-sd}, \ref{fig:compare-beta}, and \ref{fig:compare-beta-se}).
This is the first correction of summary statistics we try in the simulations below. 
Note that we recompute INFO scores for the subset of 362,307 European individuals used in this paper as they can differ substantially from the ones reported by the UK Biobank for the whole data (Figure \ref{fig:compare-info}).
As a second option, instead of using dosages to perform the GWAS, it has been argued that using multiple imputation (MI) would be more appropriate \cite[]{Palmer2016}.
In simulations, we show that $Z_j^{MI} \approx Z_j^{imp} \cdot \text{INFO}_j$ and $\hat{\gamma}_j^{MI} \approx \hat{\gamma}_j^{imp} \cdot \text{INFO}_j$ (Figure \ref{fig:compare-mi}).
This is the second correction of summary statistics we try in the simulations below, along with $n_j = N \cdot \text{INFO}_j$. 
Finally, we try an in-between solution as a third correction, using $\hat{\gamma}_j^{imp} \cdot \text{INFO}_j$, $\text{se}(\hat{\gamma}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$, and $N \cdot \text{INFO}_j$.

@@

\begin{figure}[h]
	\centerline{\includegraphics[width=0.95\textwidth]{simu-info}}
	\caption{Results for the simulations using dosage data. Correction ``sqrt\_info'' corresponds to using $\hat{\gamma}_j^{imp} \cdot \sqrt{\text{INFO}_j}$ and $\text{se}(\hat{\gamma}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$.
	Correction ``info'' corresponds to using $\hat{\gamma}_j^{imp} \cdot \text{INFO}_j$  and $N \cdot \text{INFO}_j$.}
	\label{fig:simu-info}
\end{figure}

@@

Therefore, having the INFO score enables both to correct for it in the QC step to be able to pinpoint possible other problems, and possibly to correct for it and improve prediction as shown in the simulations before.
We try this for the two breast cancer summary statistics. We first compare the standard QC proposed in \cite{prive2020ldpred2} (which ends up filtering on MAF here, which we call ``qc1''). We then also filter out the two long-range LD regions of chromosome 6 and 8 for the OncoArray summary statistics and remove around 500 variants when filtering on differences of MAFs between summary statistics and the validation dataset.
As for the correction for the INFO scores, we use $\hat{\gamma}_j^{imp} \cdot \text{INFO}_j$ and $N \cdot \text{INFO}_j$ as we have shown to perform well in simulations.
Although results for both QC used are very similar, correcting for the INFO score really improves prediction, especially for LDpred2-auto [TODO: NUMBERS] (Figure \ref{fig:res_brca}).

@@

Among genetic variants on chromosome 22 and with a minor allele frequency larger than 0.01 and an imputation INFO score larger than 0.4 (as reported by the UK Biobank), we sample 40,000 of them according to the inverse of the INFO score density so that they have varying levels of imputation accuracy (Figure \ref{fig:hist-info}).
We read the UK Biobank data into two different datasets using function \texttt{snp\_readBGEN} from R package bigsnpr \cite[]{prive2017efficient}, one by reading the BGEN data at random according to imputation probabilities, and another one reading it as dosages (i.e.\ expected values according to imputation probabilities).
The first dataset is used as what could be the real genotype calls and the second dataset as what would be its imputed version; this design technique was used in \cite{prive2019making}.

@@

We also use the UK Biobank data, and use the same individuals as described in the previous subsection. We sample 10,000 individuals to form a validation set and use the remaining 352,307 individuals as test set.
We restrict to the genetic variants to the 1,054,315 HapMap3 variants used in the LD reference provided in \cite{prive2020ldpred2}.

[TODO: MAYBE THE RIGHT PLACE TO PRESENT THE TWO NEW SETS]

@@

In this paper, we extensively use the following formula
\begin{equation}\label{eq:approx-sd-lin}
\text{sd}(G_j) \approx \dfrac{\text{sd}(y)}{\sqrt{n_j ~ \text{se}(\hat{\gamma}_j)^2 + \hat{\gamma}_j^2}} ~,
\end{equation}
where $\hat{\gamma}_j$ is the marginal (GWAS) effect of variant $j$, $n_j$ is the GWAS sample size associated with variant $j$, $y$ is the vector of phenotypes and ${G}_j$ is the vector of genotypes for variant $j$.
This formula is used in LDpred2 \cite{prive2020ldpred2,prive2021high}. Note that, for a binary trait for which logistic regression is used, we have instead
\begin{equation}\label{eq:approx-sd-log}
\text{sd}(G_j) \approx \dfrac{2}{\sqrt{n_j^\text{eff} ~ \text{se}(\hat{\gamma}_j)^2 + \hat{\gamma}_j^2}} ~,
\end{equation}
where $n_j^\text{eff} = \frac{4}{1 / n_j^\text{cases} + 1 / n_j^\text{controls}}$.

We can then impute $n_j$ from equation \eqref{eq:approx-sd-lin} using
\begin{equation}\label{eq:imputeN}
n_j \approx \dfrac{\text{var}(y) / \text{var}(G_j) - \hat{\gamma}_j^2}{\text{se}(\hat{\gamma}_j)^2} ~,
\end{equation}
and impute $n_j^\text{eff}$ from equation \eqref{eq:approx-sd-log} using
\begin{equation}\label{eq:imputeN}
n_j^\text{eff} \approx \dfrac{4 / \text{var}(G_j) - \hat{\gamma}_j^2}{\text{se}(\hat{\gamma}_j)^2} ~.
\end{equation}
In practice, we also bound this estimate to be between $0.5 \cdot N$ and $1.1 \cdot N$, where $N$ is the global sample size.

@@

Instead of using a regularized version of the correlation matrix $R$ parameterized by $s$, $R_s = (1 - s) R + s I$, we use $R_{\delta} = R + \delta I$, which makes it clearer that lassosum is also using L2-regularization (therefore elastic-net). 
Then, from \cite{mak2017polygenic}, the solution from lassosum can be obtained by iteratively updating 
\[
\beta_j^{(t)} =
\begin{cases}
\text{sign}\left(u_j^{(t)}\right) \left(\left|u_j^{(t)}\right| - \lambda\right) / \left(\widetilde{X}_j^T \widetilde{X}_j + \delta\right) & \text{if } \left|u_j^{(t)}\right| > \lambda ~, \\
0 & \text{otherwise.}
\end{cases}
\]
where 
\[
u_j^{(t)} = r_j - \widetilde{X}_j^T \left( \widetilde{X} \beta^{(t-1)} - \widetilde{X}_j \beta_j^{(t-1)} \right) ~.
\]
Following notations from \cite{prive2020ldpred2} and denoting $\widetilde{X} = \frac{1}{\sqrt{n-1}} C_n G S^{-1}$, where $G$ is the genotype matrix, $C_n$ is the centering matrix and $S$ is the diagonal matrix of standard deviations of the columns of $G$.
Then $\widetilde{X}_j^T \widetilde{X} = R_{j,.} = R_{.,j}^T$ and $\widetilde{X}_j^T \widetilde{X}_j = 1$. Moreover, $u_j^{(t)} = \beta_j^{(t-1)} + \widehat{\beta}_j - R_{.,j}^T \beta^{(t-1)}$,
where $r_j = \widehat{\beta}_j =  \dfrac{\widehat{\gamma}_j}{\sqrt{n_j ~ \text{se}(\widehat{\gamma}_j)^2 + \widehat{\gamma}_j^2}}$ and $\widehat{\gamma}_j$ is the GWAS effect of variant $j$ and $n$ is the GWAS sample size \cite[]{mak2017polygenic,prive2021high}.
Then the most time-consuming part is computing $R_{.,j}^T \beta^{(t-1)}$.
To make this faster, instead of computing $R_{.,j}^T \beta^{(t-1)}$ at each iteration ($j$ and $t$), we can start with a vector with only 0s initially (for all $j$) since $\beta^{(0)} \equiv 0$, and then updating this vector when $\beta_j^{(t)} \neq \beta_j^{(t-1)}$ only. Note that only positions $k$ for which $R_{k,j} \neq 0$ must be updated in this vector $R_{.,j}^T \beta^{(t-1)}$.

In this new implementation of the lassosum model, the input parameters are the correlation matrix $R$, the GWAS summary statistics ($\widehat{\gamma}_j$, $\text{se}(\widehat{\gamma}_j)$ and $n_j$), and the two hyper-parameters $\lambda$ and $\delta$. 
Therefore, except for the hyper-parameters, this is the exact same input as for LDpred2 \cite[]{prive2020ldpred2}.
We try $\delta \in \{0.001, 0.005, 0.02, 0.1, 0.6, 3\}$ by default in lassosum2, instead of $s \in \{0.2, 0.5, 0.8, 1.0\}$ in lassosum.
For $\lambda$, while the default in lassosum uses a sequence of 20 values equally spaced on a log scale between 0.1 and 0.001; we now use a sequence between $\lambda_0$ and $\lambda_0 / 100$ by default in lassosum2, where $\lambda_0 = \max_j \left|\widehat{\beta}_j\right|$ is the minimum $\lambda$ for which no variable enters the model because the L1-regularization is too strong.
Note that we do not provide an ``auto'' version using pseudo-validation (as in \cite{mak2017polygenic}) as we have not found it to be very robust (Figure \ref{fig:pseudoval}).
Also note that, as in LDpred2, we run lassosum2 genome-wide using a sparse correlation matrix which assumes that variants further away than 3 cM are not correlated, and therefore we do not require splitting the genome into independent LD blocks anymore (as done in lassosum).

@@

Authors thank Matthew Stephens for helpful discussions. 
Authors also thank GenomeDK and Aarhus University for providing computational resources and support that contributed to these research results.
This research has been conducted using the UK Biobank Resource under Application Number 58024.
